### 2023-10-29 paper/main.tex
lots of pointless content

\frenchspacing
\begin{document}

\section*{\raggedright
Another formula for the area of a triangle: Useless, but explicitly in map-reduce format}
\noindent \textbf{David W. Hogg} and \textbf{Soledad Villar}

\paragraph{Abstract:}
It is known that any permutation-invariant function of data points $x_i$
can be written in the form $\rho(\sum_i\phi(x_i))$,
where $\rho$ and $\phi$ are (probably nonlinear) functions, and the function $\phi$ might output high-dimensional output (but can't depend on $i$).
This form can be considered map--reduce, because the $\phi$ executions can be performed asynchronously and the sums can be performed hierarchically.
The area of a triangle is a permutation-invariant function of the locations of the three corners.
We find (embarrassingly, by symbolic regression) the formula for the area of the triangle that is explicitly in this map--reduce form.
It is ugly.

\section{Introduction}
The area $A$ of a triangle can be written in many ways:
\begin{align}
A &= \frac{1}{2}\,\text{(base)}\,\text{(height)} \\
  &= \frac{1}{2}\,|(\vec{x}_3 - \vec{x}_1) \times (\vec{x}_2 - \vec{x}_1)| \\
  &= \frac{1}{2}\,a\,b\,\sin(C) \\
  &= \sqrt{s\,(s-a)\,(s-b)\,(s-c)} \label{eq:Heron} ~,
\end{align}
where the first formula is the usual introduction to the subject,
$(\vec{x}_1, \vec{x}_2, \vec{x}_3$ are three vectors pointing from an origin to the corners of the triangle,
$\times$ represents the cross-product operator (or its generalization to higher dimensions),
$a, b, c$ are side lengths of the triangle,
$C$ is the angle between the $a$ and $b$ sides,
and $s$ is the semi-perimeter $(a + b + c)/2$.

While the area of a triangle does not depend on how its sides or corners are labeled,
only the last of these formulae---Heron's formula \eqref{eq:Heron}---is explicitly permutation-invariant.
The theory of permutation-invariant functions says (CITE) that any per\-mu\-tation-invariant function (under conditions??) of $N$ objects $x_i$ (with $1\leq i\leq N$) can be approximated (to arbitrary precision??) with an expression of the form
\begin{equation}
    \rho(\sum_{i=1}^N\phi(x_i)) \label{eq:DeepSets}
\end{equation}
where $\rho$ and $\phi$ are (possibly very nonlinear) functions,
and the $x_i$ are the data points that can be permuted without changing the result.
This result---which we associate with the subject \emph{Deep Sets} (CITE)---is not at all obvious:
For example, even though Heron's formula \eqref{eq:Heron} is permutation-invariant, it is not actually in the form \eqref{eq:DeepSets}, if the triangle is seen as being defined by its corners (or even its sides).

@@

\section{A note on our method}

\section{Result and discussion}

\begin{align}
A^2 = &
-\frac{ 54}{368}\,(\sum_{i=1}^3 x_i^2\,y_i^2) 
-\frac{258}{368}\,(\sum_{i=1}^3 x_i\,y_i)^2 \nonumber \\ &
+\frac{285}{368}\,(\sum_{i=1}^3 x_i^2)\,(\sum_{i=1}^3 y_i^2) 
+\frac{ 36}{368}\,(\sum_{i=1}^3 y_i)\,(\sum_{i=1}^3 x_i^2\,y_i) \nonumber \\ &
+\frac{ 36}{368}\,(\sum_{i=1}^3 x_i)\,(\sum_{i=1}^3 x_i\,y_i^2) 
-\frac{101}{368}\,(\sum_{i=1}^3 y_i)^2\,(\sum_{i=1}^3 x_i^2) \nonumber \\ &
+\frac{148}{368}\,(\sum_{i=1}^3 x_i)\,(\sum_{i=1}^3 y_i)\,(\sum_{i=1}^3 x_i\,y_i) 
-\frac{101}{368}\,(\sum_{i=1}^3 x_i)^2\,(\sum_{i=1}^3 y_i^2) \nonumber \\ &
+\frac{  9}{368}\,(\sum_{i=1}^3 x_i)^2\,(\sum_{i=1}^3 y_i)^2 
\end{align}

>>>>

\newcommand{\documentname}{\textsl{Note}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\frenchspacing

\begin{document}

\section*{\raggedright
Another formula for the area of a triangle: Useless, but explicitly in Deep-Sets and map--reduce form}
\noindent \textbf{David W. Hogg} and \textbf{Soledad Villar}

\paragraph{Abstract:}
It is known that any permutation-invariant function of data points $x_i$
can be written in the form $\rho(\sum_i\phi(x_i))$,
where $\rho$ and $\phi$ are (probably nonlinear) functions, and the function $\phi$ might output a high-dimensional latent encoding of each point $x_i$.
This form---known in the machine-learning literature as Deep Sets---is a map--reduce form, because the $\phi$ executions can be performed asynchronously and the sums can be performed hierarchically or in any order.
The area of a triangle is a permutation-invariant function of the locations of the three corners.
We find, possibly for the first time, the formula for the area of the triangle that is explicitly in the Deep-Sets form.
No insights of any kind were gained by this work.

\section{Introduction}
The area $A$ of a triangle can be written in many ways:
\begin{align}
A &= \frac{1}{2}\,\text{(base)}\,\text{(height)} \label{eq:school} \\
  &= \frac{1}{2}\,|(\vec{x}_3 - \vec{x}_1) \times (\vec{x}_2 - \vec{x}_1)| \label{eq:cross} \\
  &= \frac{1}{2}\,a\,b\,\sin(C) \label{eq:sine} \\
  &= \sqrt{s\,(s-a)\,(s-b)\,(s-c)} \label{eq:Heron} ~,
\end{align}
where the first formula is the usual introduction to the subject,
$(\vec{x}_1, \vec{x}_2, \vec{x}_3$ are three vectors pointing from an origin to the corners of the triangle,
$\times$ represents the cross-product operator (or its generalization to higher dimensions),
$a, b, c$ are side lengths of the triangle,
$C$ is the angle between the $a$ and $b$ sides,
and $s$ is the semi-perimeter $(a + b + c)/2$.
Below, we present a new formula for the area of a triangle, which we believe may be novel.

While the area of a triangle does not depend on how its sides or corners are labeled,
only the last of these formulae---Heron's formula \eqref{eq:Heron}---is explicitly permutation-invariant.

The theory of permutation-invariant functions says (CITE) that any per\-mu\-tation-invariant function (under conditions??) of $N$ objects $x_i$ (with $1\leq i\leq N$) can be approximated (to arbitrary precision??) with an expression of the form
\begin{equation}
    \rho(\sum_{i=1}^N\phi(x_i)) \label{eq:DeepSets}
\end{equation}
where $\rho$ and $\phi$ are (possibly very nonlinear) functions,
and the $x_i$ are the data points that can be permuted without changing the result.
This result---which we associate with the subject \emph{Deep Sets} (CITE)---is not at all obvious:
For example, even though Heron's formula \eqref{eq:Heron} is permutation-invariant, it is not actually in the form \eqref{eq:DeepSets}, if the triangle is seen as being defined by its corners (or even its sides).

@@

Although the contribution of this paper is minimal, or even zero, it was inspired by real problems in physics.
For one very specific example, all known formulae for the gravitational potential energy of a system of $N$ point masses involve sums over pairs of points; they have not ever been written in the Deep Sets form \eqref{eq:DeepSets}.
In cosmology, all known estimators for $k$-point functions (correlation functions, bispectra, and so on) involve sums over $k$-tuples; once again they have not been written in the Deep Sets form.
Importantly, these kinds of problems cannot currently be performed (not even approximately) in any map--reduce framework; indeed they are very time-consuming because they involve sums over all elements of $N$-choose-$k$, which scales (badly) as $N^k$.
Because 3-point functions involve the shapes and areas of triangles, consideration of the 3-point function led to the question answered in this \documentname:
\emph{What is the Deep Sets (or map--reduce) expression for the area of a triangle?}
We deliver in \secref{sec:result}.

\section{A note on our method}\label{sec:method}
What made us think $A^2$ rather than $A$?

What do units tell you?

How to construct all the invariant polynomials.

Which polynomial terms are invariant to other kinds of transformations?

Symbolic regression, determination of the magic number, and checking.

\section{Result and discussion}\label{sec:result}
Our result---and only contribution---is that the area $A$ of a triangle is given by
\begin{align}
A^2 = &
-\frac{ 54}{368}\,(\sum_{i=1}^3 x_i^2\,y_i^2) 
-\frac{258}{368}\,(\sum_{i=1}^3 x_i\,y_i)^2 \nonumber \\ &
+\frac{285}{368}\,(\sum_{i=1}^3 x_i^2)\,(\sum_{i=1}^3 y_i^2) 
+\frac{ 36}{368}\,(\sum_{i=1}^3 y_i)\,(\sum_{i=1}^3 x_i^2\,y_i) \nonumber \\ &
+\frac{ 36}{368}\,(\sum_{i=1}^3 x_i)\,(\sum_{i=1}^3 x_i\,y_i^2) 
-\frac{101}{368}\,(\sum_{i=1}^3 y_i)^2\,(\sum_{i=1}^3 x_i^2) \nonumber \\ &
+\frac{148}{368}\,(\sum_{i=1}^3 x_i)\,(\sum_{i=1}^3 y_i)\,(\sum_{i=1}^3 x_i\,y_i) 
-\frac{101}{368}\,(\sum_{i=1}^3 x_i)^2\,(\sum_{i=1}^3 y_i^2) \nonumber \\ &
+\frac{  9}{368}\,(\sum_{i=1}^3 x_i)^2\,(\sum_{i=1}^3 y_i)^2 ~, \label{eq:result}
\end{align}
where the two-dimensional position of each of the three corners $i$ is $(x_i, y_i)$.
This answer is in Deep-Sets form:
Every expression in parentheses in \eqref{eq:result} is a pure sum over all three points.
There are no sums over differences, and there is no ordering of the points.
Being very explicit, this can be written in the Deep Sets form \eqref{eq:DeepSets} if we define $\rho$ and $\phi$ as follows:

Note the symmetries....! See the $36/368$ terms, or the $-101/368$ terms

But note that it isn't obviously invariant to arbitrary rotations.
And it isn't obviously invariant to translations.
Formulae \eqref{eq:cross}, \eqref{eq:sine}, and \eqref{eq:Heron} are obviously rotation-invariant and translation-invariant.

Why not simplify the prefactors?

What's the generalization to $d$ dimensions?

Where does the factor of 23 come from?

What happens if we try to do $A$ as a quadratic instead of $A^2$ as a quartic?

Why not get this with computer algebra?
